---
title: "p8105_hw3_lx2346"
author: "linshan"
date: "2024-10-07"
output: github_document
---

Before start, prepare all the packages will be used
```{r}
library(ggplot2)
library(tidyverse)
```

## Problem 1
Load the data from the p8105.datasets packages
```{r}
library(p8105.datasets)
data("ny_noaa")
```

### Do some exploration of the dataset.
**size and structure**: The dataset consists of `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Each row represents a daily observation for a given weather station from `r min(ny_noaa$date)` to `r max(ny_noaa$date)`.\
**Key Variables**: The variables in the dataset are `r names(ny_noaa)`. **"id"** represents the unique identifier for each weather station. **"date"** is the date of observation. **"prep"** is precipitation(tenths of mm). **"snow"** is the snowfall(mm), while **"snwd"** is the snow depth(mm). **"tmax"** and **"tmin"** represents the maximum and minmum tempreture.\
** Missing data**: There are missing data in `r names(ny_noaa)[sapply(ny_noaa,anyNA)]`. How many missing data are shown as below. 
```{r}
sapply(ny_noaa,function(x) sum(is.na(x)))
```
So the missing value are particularly noticeable in "tmax" and "tmin".
### Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. 
separate the data variable into year, month and day.
```{r}
ny_noaa = separate(ny_noaa, date, into = c("year", "month", "day"), sep = "-", remove = FALSE)
```
 change into reasonable units
```{r}
ny_noaa = ny_noaa |>
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         prcp = as.numeric(prcp),
         snow = as.numeric(snow),
         snwd = as.numeric(snwd)) |>
  mutate(tmax = tmax/10,
         tmin = tmin/10,
         prcp = prcp/10)
```
#### For snowfall, what are the most commonly observed values? Why?
```{r}
ny_noaa |>
  count(snow) |>
  arrange(desc(n)) |>
  slice(1)
```
It shows that the value 0 are most commonly observed, which is normal, since it doesn't snow in New York for most of the year.

### Make a two-panel plot showing the average max temperature in January and in July in each station across years.
```{r}
ny_noaa |>
  filter(month == "01" | month == "07") |>
  group_by(month, id) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |>
  filter(!if_any(everything(), is.na))|>
  ggplot(aes(x = id, y = mean_tmax, color = month)) +
  geom_point() +
  facet_grid(.~ month, scale = "free_y", labeller = labeller(month = c("01" = "January", "07" = "July"))) +
  labs(
    title = " Average Maximum Temperature in January and in July",
    x = "Weather stations",
    y = "Maximum daily tempreture(C)") +
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
  
```

### Is there any observable / interpretable structure? Any outliers?
It can be observed that **July's tmax is significantly higher than January's**, with most of July's tmax centered around 30 degrees, while most of January's average temperature is centered around 0 degrees. Among the mean maximum temperatures in January, there are three “outliers” around -10 degree Celsius.

### Make a two-panel plot showing tmax vs tmin for the full dataset
```{r}
ny_noaa |>
  pivot_longer(
    tmax:tmin,
    names_to = "ob_t",
    values_to = "tempreture"
  ) |>
  drop_na() |>
  ggplot(aes(x = date, y = tempreture, fill = ob_t)) +
  geom_bin2d(bins = 200, alpha = .8, size = 0.5, na.rm = TRUE) +
  scale_y_continuous(
    breaks = c(-20, -10, 0, 10, 20 ,30, 40),
    labels = c("-20C", "-10C","0C", "10C","20C", "30C", "40C")
  ) +
  scale_fill_discrete(name = "observations") +
  facet_grid(. ~ ob_t)
```

### Make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.
```{r}
ny_noaa |>
  filter(snow > 0, snow < 100) |>
  ggplot(aes(x = year, y = snow)) + 
  geom_violin(aes(fill = year), alpha = .5) +
  scale_x_discrete(
    breaks = c(1981, 1985, 1990, 1995, 2000, 2005, 2010),
    labels = c("1981", "1985","1990", "1995","2000", "2005", "2010")
  ) +
  stat_summary(fun = "median", color = "black", size = .1)
```



























